Efficient pedigree recording for fast population genetics simulation

### 2017-12-01 forwards_paper.tex
Reorg benchmarks section; add simuPOP

\include{macros}
\newcommand{\simupop}{\texttt{simuPOP}}
\newcommand{\fwdpp}{\texttt{fwdpp}}
\newcommand{\msprime}{\texttt{msprime}}
\newcommand{\nodes}{\texttt{nodes}}
\newcommand{\edgesets}{\texttt{edgesets}}
\newcommand{\sites}{\texttt{sites}}
\newcommand{\mutations}{\texttt{mutations}}
\newcommand{\setdiff}{\smallsetminus}
\newcommand{\Nt}{\mathcal{N}}  % node table
\newcommand{\Et}{\mathcal{E}}  % edge table
\newcommand{\St}{\mathcal{S}}  % site table
\newcommand{\Mt}{\mathcal{M}}  % mutation table
\newcommand{\Al}{\mathcal{A}}  % list of ancestors
\newcommand{\taddnode}[1]{\mbox{addnode}(#1)}
\newcommand{\taddedge}[1]{\mbox{addedge}(#1)}
\newcommand{\tsimplify}[1]{\mbox{simplify}(#1)}
\DeclareMathOperator{\len}{length}
\DeclareMathOperator{\add}{add} % for adding to tables
\newcommand{\lleft}{\mathrm{left}}  % left label
\newcommand{\lright}{\mathrm{right}}
\newcommand{\lid}{\mathrm{id}}

@@

We begin by benchmarking the performance improvement achieved by connecting
two previously published forwards-time simulation libraries,
\simupop{} \citep{simupop} and \fwdpp{} \citep{fwdpp}
to the tools we describe here, implemented in \msprime{}.
Then, we describe the conceptual and algorithmic foundations for the method:
(a) a format, implemented in the \msprime{} Python API,
for recording tree sequences in several \emph{tables};
(b) an algorithm, \algref{W}, to record tree sequence information into these tables on the fly
    from a forwards-time simulation;
and (c) an algorithm to \emph{simplify} a tree sequence, i.e., remove redundant information.
We then describe a general-purpose method that uses these tools
(as implemented in \msprime{})
to efficiently record simplified a tree sequence,
and analyze of its run time complexity.

@@

To measure the performance gains from ancestry recording, we ran simulations implemented in C++ (see above) with and
without ancestry tracking.  The simulations without ancestry tracking introduced neutral mutations at a scaled mutation
rate equal to the scaled recombination rate (\textit{e.g.}, $\theta = 4N\mu = \rho$, where $\rho = 4Nr$, $N$ is the
diploid population size, $\mu$ is the
mutation rate to neutral mutations (per gamete, per generation) and $r$ is the recombination rate per diploid per
generation.  Both $\mu$ and $r$ are the means of a Poisson process and we used the infinitely-many sites mutation model
\citep{Kimura1969-uc}.

We varied $N \in (1000, 10000, 50000)$, $\rho \in (10^3, 10^4, 10^5)$ and ran a single replicate of all $(N,\rho)$
combinations.  Deleterious mutations were introduced at rate $\rho/100$ per generation, drawing scaled selection
coefficients ($2Ns$)
from a Gamma distribution with a mean of -5 and a shape parameter of 1.  This distribution of fitness effects results in
thousands of weakly-deleterious mutations segregating in the population, many of which drift to intermediate
frequencies.

We ran all benchmarks on an Ubuntu Linux (version 16.04) system with two 2.6 Gigahertz Intel E5-2650 CPU with
hyperthreading enabled.  We ran one simulation at a time and the machine was under minimal load otherwise.  We used GNU
parallel \citep{Tange2011a} to kill
any simulation that did not finish within 72 hours.  We used the Linux \texttt{time} command to record run time and peak
memory usage of each replicate and our simulation scripts separately recorded the time spent in various steps of each
run.

For efficiency reasons, we only ran \simupop{} with ancestry tracking.

\krt{Jaime--any details here to add re: how simuPOP was tweaked??}

% \plr{Jaime and Kevin to describe simulation methods here (separately).}
% 
% Comparison of simulation with/without msprime, using \simupop{}
% or maybe just a simple haploid simulation with 1000 QTL and stabilizing selection on a trait (say).
% 
% Maybe an estimate of how long \emph{just} the ARG recording and simplification takes,
% so that then we can say how fast the simulator would have to be to do $10^6$ whole chromosomes for $10^7$ generations
% in a day.

>>>>

\include{macros}
\newcommand{\simupop}{\texttt{simuPOP}}
\newcommand{\fwdpp}{\texttt{fwdpp}}
\newcommand{\msprime}{\texttt{msprime}}
\newcommand{\ftprime}{\texttt{ftprime}}
\newcommand{\nodes}{\texttt{nodes}}
\newcommand{\edgesets}{\texttt{edgesets}}
\newcommand{\sites}{\texttt{sites}}
\newcommand{\mutations}{\texttt{mutations}}
\newcommand{\setdiff}{\smallsetminus}
\newcommand{\Nt}{\mathcal{N}}  % node table
\newcommand{\Et}{\mathcal{E}}  % edge table
\newcommand{\St}{\mathcal{S}}  % site table
\newcommand{\Mt}{\mathcal{M}}  % mutation table
\newcommand{\Al}{\mathcal{A}}  % list of ancestors
\newcommand{\taddnode}[1]{\mbox{addnode}(#1)}
\newcommand{\taddedge}[1]{\mbox{addedge}(#1)}
\newcommand{\tsimplify}[1]{\mbox{simplify}(#1)}
\DeclareMathOperator{\len}{length}
\DeclareMathOperator{\add}{add} % for adding to tables
\newcommand{\lleft}{\mathrm{left}}  % left label
\newcommand{\lright}{\mathrm{right}}
\newcommand{\lid}{\mathrm{id}}

@@

We begin by benchmarking the performance improvement achieved by connecting
two previously published forwards-time simulation libraries,
\simupop{} \citep{simupop} and \fwdpp{} \citep{fwdpp}
to the tools we describe here, implemented in \msprime{}, allowing them to
efficiently record the population pedigree in a simplified tree sequence.
Then, we describe the conceptual and algorithmic foundations for the method:
(a) a format, implemented in the \msprime{} Python API,
for recording tree sequences in several \emph{tables};
(b) an algorithm, \algref{W}, to record tree sequence information into these tables on the fly
    from a forwards-time simulation;
and (c) an algorithm to \emph{simplify} a tree sequence, i.e., remove redundant information.
Finally, we analyze the run time and space complexity of our general-purpose method.

@@

To measure the performance gains from recording the pedigree we ran simulations with and
without recording, for both implementations (\simupop{} \citep{simupop}
and \fwdpp{} \citep{fwdpp}).
All simulations implemented a discrete-time Wright-Fisher population of diploid
population size $N$, simulated for $10N$ generations.
The simulations without pedigree recording introduced neutral mutations at a scaled mutation
rate equal to the scaled recombination rate (\textit{e.g.}, $\theta = 4N\mu = \rho$, where $\rho = 4Nr$, $\mu$ is the
mutation rate to neutral mutations (per gamete, per generation) and $r$ is the recombination rate per diploid per
generation.
Deleterious mutations were introduced at rate $\rho/100$ per generation, drawing scaled selection
coefficients ($2Ns$)
from a Gamma distribution with a mean of -5 and a shape parameter of 1.  This distribution of fitness effects results in
thousands of weakly-deleterious mutations segregating in the population, many of which drift to intermediate
frequencies.  The case of many mutations with selection is a non-trivial violation of exchangeability assumptions of the
coalescent \citep{Neuhauser1997-nn}.  Therefore, these selected mutations must be explicitly tracked in our forward simulation
and the time savings due to pedigree recording come from not having to record \textit{neutral} mutations.

We ran all benchmarks on an Ubuntu Linux (version 16.04) system with two 2.6 Gigahertz Intel E5-2650 CPU with
hyperthreading enabled.  We ran one simulation at a time and the machine was under minimal load otherwise.  We used GNU
parallel \citep{Tange2011a} to kill
any simulation that did not finish within 72 hours.  We used the Linux \texttt{time} command to record run time and peak
memory usage of each replicate and our simulation scripts separately recorded the time spent in various steps of each
run.
We varied $N$ and $\rho$ and ran a single replicate of all $(N,\rho)$
combinations (the actual values differed for \simupop{} and \fwdpp{}, see below).

\subsubsection*{The \fwdpp{} implementation}

We implemented Wright-Fisher simulations in C++ using \fwdpp{} library
functions and interface code that records data into the \msprime{} Tables API (see below).
In these simulations, $\mu$ and $r$ are the means of a Poisson process and we used the infinitely-many sites mutation model
\citep{Kimura1969-uc}.
We varied $N \in (1000, 10000, 50000)$, $\rho \in (10^3, 10^4, 10^5)$.

\subsubsection*{The \simupop{} implementation}

We implemented Wright-Fisher simulations in Python using \simupop{} library
functions (which wrap C++ code) and interface code (\ftprime{}, also in Python) that records data into the \msprime{} Tables API (see below).
In these simulations, $\mu$ and $r$ define a per-base-pair mutation rate and we
defined the length of the simulated sequence, in base pairs, as $L = \tfrac{\rho}{4N \mu}$.
We varied $N \in (1000, 10000)$, $\rho \in (10^3, 10^4)$.

% Comparison of simulation with/without msprime, using \simupop{}
% or maybe just a simple haploid simulation with 1000 QTL and stabilizing selection on a trait (say).
% 
% Maybe an estimate of how long \emph{just} the ARG recording and simplification takes,
% so that then we can say how fast the simulator would have to be to do $10^6$ whole chromosomes for $10^7$ generations
% in a day.
